---
title: "Assignment 1"
author: "Kyle Sorstokke"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(tidyverse)
```


```{r dataEntry, echo=FALSE}
deaths <- read.csv("./data/deaths.csv")
ylds <- read.csv("./data/ylds.csv")
l_e <- read.csv("./data/life_expectancy.csv")
```

## Describing Metrics

  * YLLs: Years of Life Lost are a scaled measurement of deaths, where each
  death is counted by the years lost according to life expectancy of that person's
  age group. They represent the total loss of life from a certain cause, and they
  are calculated by multiplying the number of deaths in each age group by that age's
  life expectancy.
  * YLDs: Years Lost to Disability is a measurement of the total healthy years
  lost to a certain cause. It is calculated by scaling the prevalence (number of cases)
  by the disability weight for each year the person is expected to live with the disability.
  * DALYs: Disability adjusted life years, calculated as the sum of YLLs and
  YLDs, measure the total impact of a disability.
  * Paper: In the paper given, disability weights are calculated on a scale
  of 0 to 1 through comparison of different disabilities. By scaling each against
  the others, they are able to effectively estimate the overall health impact of each.
  * Age: Age discounting reduces emphasis on future impact in favor of immediate
  (present) impact by reducing the multiplier on future years. Age weighting places
  higher emphasis on young adult years because of their perceived contributions
  to society in their "prime". I don't believe that age weighting should be included
  in DALYs; as researchers measuring health impact of disabilities, it isn't our
  job to determine which years of an individual's life have the most value or
  weight. However, I do believe in focusing on issues that face us in the immediate
  future, so I think there is a place for age discounting in certain calcuations
  of DALYs.
\end{itemize}

## Computing Metrics

```{r computing, echo=FALSE}
deaths <- deaths %>%
  left_join(l_e, by = "age") %>%
  replace(is.na(.),0)
deaths <- deaths %>%
  mutate(yll = deaths * life.expectancy)
deaths <- deaths %>%
  left_join(ylds, by = c("cause_name", "sex", "age")) %>%
  replace(is.na(.),0)
deaths <- deaths %>%
  mutate(daly = yll + ylds)
```


## Burden By Cause

```{r cause, warning = F, message = F}
by_cause <- deaths %>%
  group_by(cause_name) %>%
  summarize(tDALY = sum(daly), tYLL = sum(yll), tYLD = sum(ylds))

plot(by_cause$tYLL, by_cause$tYLD, xlab = "YLLs", ylab = "YLDs")
```

The plot above shows that there is somewhat of an inverse relationship between YLLs
and YLDs, which makes sense intuitively; disabilities with high mortalities won't
have people live as long with them and vice verse. The DALYs show a more holistic
total impact that doesn't show the difference between parts.

## Burden By Age

```{r age, warning = F, message = F}
by_age <- deaths %>%
  group_by(age) %>%
  summarize(tDALY = sum(daly), tYLL = sum(yll), tYLD = sum(ylds)) %>%
  pivot_longer(cols = c(tYLL, tYLD))

ggplot(by_age, aes(fill=name, y=value, x=age)) + 
    geom_bar(position="stack", stat = "identity") +
  theme(axis.text.x=element_text(angle = 90, hjust = 0))
```

The stacked barplot above shows that the Under 5 age group has by far the biggest
aggregate years of life lost, which makes sense because they have the highest life
expectancy and many conditions studied hit this age group relatively hard. We can
also see that YLLs (and YLDs) decline with life expectancy for older groups. The
overall trend in DALYs is that infants are at high risk, then risk is relatively
low until adulthood when more disabilities start to come into play.


## By Sex

```{r sex, warning = F, message = F}
by_sex <- deaths %>%
  group_by(sex) %>%
  summarize(tDALY = sum(daly), tYLL = sum(yll), tYLD = sum(ylds)) %>%
  pivot_longer(cols = c(tYLL, tYLD, tDALY))


ggplot(by_sex, aes(fill=sex, y=value, x=name)) + 
    geom_bar(position=position_dodge(), stat = "identity")
```

The chart above shows that males face a higher total impact from DALYs, as well as
a noticeably higher total years of life lost. However, we can see that females
have a slightly higher total years lived with disability. This would suggest that
males tend more towards disabilities with higher mortality rates than women do.









